Synthesising Termination Proofs for Bit-Vector Programs

### 2014-07-06 papers/termination/paper.tex
Edited introduction

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

We present a reduction from program \emph{termination} to program
\emph{synthesis}.  This reduction allows us to generate nonlinear,
lexicographic ranking functions that are correct for fixed-width machine
arithmetic and floating-point arithmetic.  We use this reduction to build a
sound and complete procedure for the termination of fixed-width and
floating-point arithmetic programs.
%
\end{abstract}

@@

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating-Point Ranking Functions.

@@

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.  a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program terminates.

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
Briefly, the transition relation is shown to be disjunctively well founded
by finding a set of simple functions, each of which proves well foundedness
for part of the transition relation.  The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require non-linear
functions or lexicographic orders.

@@

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.

@@

By assuming bit-vector semantics to be identical to integer semantics, these
techniques ignore the wrap-around behaviour caused by overflows, which can
be unsound.  In Section~\ref{sec:motivation}, we show that integers and
bit-vectors exhibit incomparable behaviours with respect to termination,
i.e.  programs that terminate for integers may \emph{not} terminate for
bit-vectors and vice versa.  Thus, abstracting bit-vectors to integers may
give rise to {\em unsound} and {\em incomplete} analyses.

We propose a general framework that uniformly computes lexicographic,
non-linear ranking functions supported by non-linear inductive invariants
for loops with arbitrary guards and transitions over bit-vectors and floats. 
It also computes non-linear recurrence sets, which comprise non-termination
proofs.  Our encoding treats multiple loops and nested loops uniformly,
without having to enumerate lassos.

 For this purpose, we phrase the termination and non-termination problems as
second-order satisfaction problems.  We identify a fragment of second-order
logic that is expressive enough to capture our formulation, which we call
the \emph{synthesis fragment}.  Subsequently, we propose a method for
checking the satisfiability of a formula in the synthesis fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
if a pair of a ranking function and a supporting invariant exist it will be
found, but if the program does not terminate we may not find a recurrence
set.  However, if the program's transition relation is deterministic, our
method is complete.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.

This means that our procedure is not directly dependent on the structure of
the analysed loop (e.g.~the number of variables or the number of lassos),
but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
We argue that the length of a program's termination proof is indicative of
how easy to understand the program is, and thus programmers tend to write
programs that have relatively short termination proofs.  We empirically show
that this claim holds in practice, causing our technique to perform well.

@@

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

\item  We rephrased the termination and non-termination problems as
second-order satisfaction problems and used program synthesis techniques to
solve them.  In contrast to previous approaches to program synthesis, we
found that using genetic programming was very effective at generating
candidate termination and non-termination proofs.

\item Our procedure does not require the enumeration of lassos, or directly
reasoning about the transitive closure of the transition relation.  To
verify our proofs, we only require a single call to a SAT solver.

\item Our procedure is based on a series of source-to-source
transformations.  As well as leading to a concise implementation, this has
the side effect of ensuring that our proof is correct for the exact binary
produced by a particular compiler, rather than for an abstraction of the
program.

@@

\item We propose a non-trivial formulation of the (non-)termination problem inside a
 second-order logic fragment with only one quantifier alternation.

@@

{\bf Limitations.}
Our algorithm decides termination of transition systems with finite state
spaces.  The (non-)termination proofs take the form of ranking functions and
program invariants that are expressed in a quantifier free language.  This
formalism is powerful enough to handle a large fragment of C, but is not
rich enough to analyse code using any of the following features:

\begin{itemize}
\item Recursion (could probably do this).
\item Heap (Thor).
\item Arrays.
\end{itemize}

Additionally, if the program under analysis is non-terminating and has a
non-deterministic transition relation, our procedure is not guaranteed to find a non-termination proof.

>>>>

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
ranking function for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

We present a reduction from program \emph{termination} to program
\emph{synthesis} via a fragment of second-order logic.  This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination of fixed-width and floating-point
arithmetic programs.
%
\end{abstract}

@@

\keywords
Termination, Non-Termination,
Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating-Point Ranking Functions.

@@

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.  a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program with a fixed amount of
storage terminates.

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter proof style builds an argument that a transition relation is disjunctively well founded
by composing several small well-foundedness arguments.
The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require non-linear
functions or lexicographic orders.

@@

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.  Furthermore, our approach allows
us to simultaneously search for proofs of non-termination, which take the form
of recurrence sets.

@@

By assuming bit-vector semantics to be identical to integer semantics, these
techniques ignore the wrap-around behaviour caused by overflows, which can
be unsound.  In Section~\ref{sec:motivation}, we show that integers and
bit-vectors exhibit incomparable behaviours with respect to termination,
i.e.  programs that terminate for integers may \emph{not} terminate for
bit-vectors and vice versa.  Thus, abstracting bit-vectors with integers may
give rise to {\em unsound} and {\em incomplete} analyses.

% we propose a general framework that uniformly computes lexicographic,
% non-linear ranking functions supported by non-linear inductive invariants
% for loops with arbitrary guards and transitions over bit-vectors and floats. 
% it also computes non-linear recurrence sets, which comprise non-termination
% proofs.  our encoding treats multiple loops and nested loops uniformly,
% without having to enumerate lassos.

Our approach is to phrase the termination and non-termination problems as
second-order satisfaction.  We identify a fragment of second-order
logic that is expressive enough to capture both problems, while being
restrictive enough to admit effective solvers.  Subsequently, we propose a method for
checking the satisfiability of a formula in this fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and
complete for deterministic bit-vector programs -- for any program, we will find
either a proof of its termination or non-termination.  For non-deterministic
programs, we are only semi-complete -- if the program terminates, we will find a proof,
but if it does not terminate we may fail to discover a recurrence set.

% With respect to the performance of our technique, we show that its runtime is dominated by the size of 
% the shortest termination proof.
% 
% This means that our procedure is not directly dependent on the structure of
% the analysed loop (e.g.~the number of variables or the number of lassos),
% but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
% We argue that the length of a program's termination proof is indicative of
% how easy to understand the program is, and thus programmers tend to write
% programs that have relatively short termination proofs.  We empirically show
% that this claim holds in practice, causing our technique to perform well.

@@

\item We designed a bit-level accurate technique for computing ranking
functions and recurrence sets that correctly accounts for the wrap-around
behavior caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program simultaneously, including
nested loops.
% In contrast to previous approaches to program synthesis, we
% found that using genetic programming was very effective at generating
% candidate termination and non-termination proofs.

\item We used program synthesis techniques to solve the second-order satisfaction
problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
proof.  It is not based on enumerating lassos, nor does it require a safety prover
that can reason about transitive closure.

\item Our solver is based on a series of source-to-source
transformations.  As well as leading to a concise implementation, this has
the side effect of ensuring that our proofs are correct for the exact binary
produced by a particular compiler, rather than for an abstraction of the
program.

@@

@@

{\bf Limitations.}
Our algorithm proves termination for transition systems with finite state
spaces.  The (non-)termination proofs take the form of ranking functions and
program invariants that are expressed in a quantifier free language.  This
formalism is powerful enough to handle a large fragment of C, but is not
rich enough to analyse code which uses the heap or arrays.  Also, we have not
yet added support for recursion to our encoding.
